---
title: "Project ID: <insert Project ID>"
output: html_notebook
params:
  raw_data: "raw"
  fwd_seq: "AAACTYAAAKGAATTGRCGG"
  rev_seq: "ACGGGCGGTGWGTRC"
  fwd_pat: "R1(_001){0,1}.fastq.gz"
  rev_pat: "R2(_001){0,1}.fastq.gz"
  fname_regex: "-MEL17453A[:digit:]{1,}-AAHLCF3M5_S[:digit:]{1,}_L001_R1_001\\.fastq\\.gz"
  read_len: 300 #DO NOT QUOTE
  isits: FALSE
  cpus: 8 #DO NOT QUOTE
  cutadapt_path: "/apps/z_install_tree/linux-rocky8-ivybridge/gcc-12.2.0/cutadapt-4.3-h7up35wj3xjo6pfjkkevxbkqd764obtw/bin/cutadapt"
  cdhit_path: "/apps/z_install_tree/linux-rocky8-ivybridge/gcc-12.2.0/cdhit-4.8.1-d3fwnap7tmcfkp2h46bxgntfddriozt5/bin" #to the binaries folder not the executable
  db_train: "/srv/scratch/ferrari/utils/silva_nr99_v138.1_train_set.fa.gz"
  db_assign: ""
---

# Setup 

## Load libraries
```{r libraries}
library(pacman)

library_list <- list(
  "magrittr",
  "dada2",
  "ShortRead",
  "Biostrings",
  "tibble",
  "dplyr",
  "tidyr",
  "stringr",
  "phylotools",
  "phyloseq",
  "pacman"
)

p_load(char = library_list, install = FALSE, character.only = TRUE)

pkg_info <- lapply(library_list, packageDescription, fields = c("Package", "Version"))

do.call(rbind, pkg_info)
```


```{r files}
path <- params$raw_data  ## directory containing the fastq files.
list.files(path, recursive = F)
```

## Check Phred scores

New Illumina data uses binned Phred scores, not a standard continuous scale of values. This affects to the error rate estimation during DADA2's downstream analysis. This chunk detects the type of scores in the dataset.

```{r phred}
fq_qa <- qa(path, type = "fastq")
myscores <- fq_qa@.srlist$baseQuality %>% filter(count > 0) %>% distinct(score) %>% pull(score)
qual_list <- as(PhredQuality(myscores), "IntegerList") %>% unlist()
qual_list

if (length(qual_list) < 5) {
  phred_scores = "binned"
} else {
  phred_scores = "standard continuous"
}
```

The Phred scores in the FastQ files are: `r qual_list`. Most likely the data contains `r phred_scores` Phred scores.

# Processing

## Identify primers
```{r listFastq}
fnFs <- sort(list.files(path, pattern = params$fwd_pat, full.names = TRUE))
fnRs <- sort(list.files(path, pattern = params$rev_pat, full.names = TRUE))
```


```{r set_primers}
FWD <- params$fwd_seq  ## CHANGE ME to your forward primer sequence
REV <- params$rev_seq  ## CHANGE ME...
```


```{r orient}
allOrients <- function(primer) {
  # Create all orientations of the input sequence
  require(Biostrings)
  dna <- DNAString(primer)  # The Biostrings works w/ DNAString objects rather than character vectors
  orients <- c(Forward = dna,
               Complement = Biostrings::complement(dna),
               Reverse = Biostrings::reverse(dna), 
               RevComp = Biostrings::reverseComplement(dna))
  return(sapply(orients, toString))  # Convert back to character vector
}
FWD.orients <- allOrients(FWD)
REV.orients <- allOrients(REV)
FWD.orients
```


```{r filterAndTrim}
fnFs.filtN <- file.path(path, "filtN", basename(fnFs)) # Put N-filterd files in filtN/ subdirectory
fnRs.filtN <- file.path(path, "filtN", basename(fnRs))
filterAndTrim(fnFs, fnFs.filtN, fnRs, fnRs.filtN, maxN = 0, multithread = params$cpus) ### mc.cores added
```


```{r primerHits}
primerHits <- function(primer, fn) {
  # Counts number of reads in which the primer is found
  nhits <- vcountPattern(primer, sread(readFastq(fn)), fixed = FALSE)
  return(sum(nhits > 0))
}
rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn = fnFs.filtN[[1]]), 
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn = fnRs.filtN[[1]]), 
      REV.ForwardReads = sapply(REV.orients, primerHits, fn = fnFs.filtN[[1]]), 
      REV.ReverseReads = sapply(REV.orients, primerHits, fn = fnRs.filtN[[1]]))
```

## Remove Primers
```{r cutadapt_set}
cutadapt <- params$cutadapt_path # CHANGE ME to the cutadapt path on your machine
system2(cutadapt, args = "--version") # Run shell commands from R
```


```{r cutadapt_run}
path.cut <- file.path(path, "cutadapt")
if(!dir.exists(path.cut)) dir.create(path.cut)
fnFs.cut <- file.path(path.cut, basename(fnFs))
fnRs.cut <- file.path(path.cut, basename(fnRs))

FWD.RC <- dada2:::rc(FWD)
REV.RC <- dada2:::rc(REV)
# Trim FWD and the reverse-complement of REV off of R1 (forward reads)
R1.flags <- paste("-g", FWD, "-a", REV.RC) 
# Trim REV and the reverse-complement of FWD off of R2 (reverse reads)
R2.flags <- paste("-G", REV, "-A", FWD.RC) 
# Run Cutadapt
for(i in seq_along(fnFs)) {
  system2(cutadapt, args = c(R1.flags, R2.flags, "-n", 2, # -n 2 required to remove FWD and REV from reads
                             "-j", "0", #autodetect CPUS
                             "-o", fnFs.cut[i], "-p", fnRs.cut[i], # output files
                             "-m", "10:10", #discard any read shorter than LEN:LEN2
                             "--nextseq-trim=20", #trim polyG
                             fnFs.filtN[i], fnRs.filtN[i])) # input files
}
```


```{r primerhitsbind}
rbind(FWD.ForwardReads = sapply(FWD.orients, primerHits, fn = fnFs.cut[[1]]), 
      FWD.ReverseReads = sapply(FWD.orients, primerHits, fn = fnRs.cut[[1]]), 
      REV.ForwardReads = sapply(REV.orients, primerHits, fn = fnFs.cut[[1]]), 
      REV.ReverseReads = sapply(REV.orients, primerHits, fn = fnRs.cut[[1]]))
```

```{r sample_names}
# Forward and reverse fastq filenames have the format:
cutFs <- sort(list.files(path.cut, pattern = params$fwd_pat, full.names = TRUE))
cutRs <- sort(list.files(path.cut, pattern = params$rev_pat, full.names = TRUE))

# Extract sample names, assuming filenames have format:
get.sample.name <- function(fname) {
  stringr::str_remove(basename(fname), params$fname_regex)
}
sample.names <- get.sample.name( cutFs )

sample.names
```


## Inspect read quality profiles
```{r qcFwd}
plotQualityProfile(cutFs[1:2]) + 
  ggplot2::scale_x_continuous(breaks = seq(0, params$read_len, 25))
```


```{r qcRev}
plotQualityProfile(cutRs[1:2]) + 
  ggplot2::scale_x_continuous(breaks = seq(0, params$read_len, 25))
```


## Filter and trim
```{r listFilt}
filtFs <- file.path(path.cut, "filtered", basename(cutFs))
filtRs <- file.path(path.cut, "filtered", basename(cutRs))
```

```{r filterAndTrim_quality}
out <- filterAndTrim(cutFs, filtFs, cutRs, filtRs, 
                     maxN = 0, maxEE = c(3, 5),
                     truncLen = c(275,230), truncQ = 2, minLen = 150,
                     rm.phix = TRUE, compress = TRUE, multithread = params$cpus)  # on windows, set multithread = FALSE

out
```


## Learn the Error Rates

For binned quality scores, you need to create an alternate to the `loessErrfun` that `learnErrors` usually uses:
```{r errfun}
if (phred_scores == "binned") {
  binned_error_fun <- makeBinnedQualErrfun(qual_list)
} else {
  binned_error_fun <- loessErrfun
}
```


```{r learnErrorsF}
errF <- learnErrors(filtFs, 
                    errorEstimationFunction = binned_error_fun,
                    multithread = params$cpus)
```


```{r learnErrorsR}
errR <- learnErrors(filtRs, 
                    errorEstimationFunction = binned_error_fun,
                    multithread = params$cpus)
```


```{r plotErrorsF}
plotErrors(errF, nominalQ = TRUE)
```

```{r plotErrorsR}
plotErrors(errR, nominalQ = TRUE)
```

## Dereplicate identical reads
```{r derep}
derepFs <- derepFastq(filtFs[stringr::str_detect(filtFs, "No_Template_Control*", negate = TRUE)], verbose = TRUE)
derepRs <- derepFastq(filtRs[stringr::str_detect(filtRs, "No_Template_Control*", negate = TRUE)], verbose = TRUE)
# Name the derep-class objects by the sample names
names(derepFs) <- sample.names[stringr::str_detect(filtFs, "No_Template_Control*", negate = TRUE)]
names(derepRs) <- sample.names[stringr::str_detect(filtRs, "No_Template_Control*", negate = TRUE)]
```


## Sample Inference
```{r remove_errors}
dadaFs <- dada(derepFs, err = errF, multithread = params$cpus)
dadaRs <- dada(derepRs, err = errR, multithread = params$cpus)
```


## Merge paired reads
```{r mergePairs}
mergers <- mergePairs(dadaFs, derepFs, dadaRs, derepRs, verbose=TRUE, justConcatenate = FALSE)
rm(derepFs)
rm(derepRs)
```


## Construct Sequence Table
```{r makeSequenceTable}
seqtab <- makeSequenceTable(mergers)
dim(seqtab)
```

## Remove chimeras
```{r removeBimeraDenovo}
seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread = params$cpus, verbose=TRUE)
```

```{r chimera_prop}
sum(seqtab.nochim)/sum(seqtab)
```


```{r sizes}
table(nchar(getSequences(seqtab.nochim)))
```

```{r writeSeqtabnochim}
write.csv(seqtab.nochim, "seqtab.nochim.csv")
```


## "Manual" collapse
```{r write_nochim_fasta}
data.frame(seq.name = colnames(seqtab.nochim),
           seq.text = colnames(seqtab.nochim)) %>% 
  mutate(seq.name = paste0(">", seq.name)) %>% 
  write.table("seqtab.nochim.fasta", sep = "\n", col.names = F, row.names = F, quote = F)
```

```{r cdhit_info}
cdhitest <- paste0(params$cdhit_path, "/cd-hit-est") # CHANGE ME to the cutadapt path on your machine
clstr2txt <- paste0(params$cdhit_path, "/clstr2txt.pl")
system2(cdhitest, args = "--version") # Run shell commands from R
```


```{r cd-hit-est}
system2(cdhitest, args = c("-i", "seqtab.nochim.fasta",
                           "-o", "seqtab.nochim.collapse.fasta",
                           "-T", params$cpus,
                           "-M", "0",
                           "-c", "1",
                           "-d", "0",
                           "-g", "1"))
```


```{r clstr2txt}
system2(clstr2txt, 
        args = c("seqtab.nochim.collapse.fasta.clstr"),
        stdout = "seqtab.nochim.collapse.fasta.tsv")
```


```{r load_clstr}
collapse_clstr <- read.table("seqtab.nochim.collapse.fasta.tsv",
                             sep = "\t", header = T) %>% 
  select(id, clstr, clstr_rep)

clstr_reps <- collapse_clstr %>% 
  filter(clstr_rep == 1) %>% 
  select(-clstr_rep)
```


```{r seqtab_clstr}
seqtab.clstr <- seqtab.nochim %>% 
  as.data.frame() %>% 
  tibble::rownames_to_column(var = "sample") %>%
  tidyr::pivot_longer(cols = -sample, names_to = "ASV", values_to = "counts") %>% 
  left_join(collapse_clstr, by = c("ASV" = "id"), multiple = "all") %>% 
  select(sample, clstr, counts) %>% 
  group_by(sample, clstr) %>% 
  summarise(sum = sum(counts)) %>% 
  ungroup()
```

```{r seqtab_nochim_collapse}
seqtab.nochim.collapse <- seqtab.clstr %>% 
  left_join(clstr_reps, by = "clstr") %>% 
  select(sample, id, sum) %>% 
  tidyr::pivot_wider(names_from = id, values_from = sum, values_fill = 0, names_repair = "minimal") %>% 
  tibble::column_to_rownames(var = "sample") %>% 
  data.matrix()
```


## Track reads through the pipeline
```{r trackReads}
require(magrittr)
require(dplyr)
require(tibble)

getN <- function(x) sum(getUniques(x))
out2 <- out
rownames(out2) <- unname(sapply(rownames(out2), get.sample.name))

track <- tibble::as_tibble(out2, rownames = "Sample") %>% 
  full_join(tibble::as_tibble(sapply(dadaFs, getN), rownames = "Sample"), by = "Sample") %>% 
  full_join(tibble::as_tibble(sapply(dadaRs, getN), rownames = "Sample"), by = "Sample") %>% 
  full_join(tibble::as_tibble(sapply(mergers, getN), rownames = "Sample"), by = "Sample") %>% 
  full_join(tibble::as_tibble(rowSums(seqtab.nochim), rownames = "Sample"), by = "Sample") %>% 
  full_join(tibble::as_tibble(rowSums(seqtab.nochim.collapse), rownames = "Sample"), by = "Sample")
# If processing a single sample, remove the sapply calls: e.g. replace
# sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("Sample", "input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim", "collapsed")
#rownames(track) <- sample.names
head(track)
track

as.data.frame(track) %>%
  tibble::rownames_to_column(var = "sample") %>%
  select(sample, everything()) %>%
  write.table("read_tracking.tsv", row.names = F, sep = "\t", col.names = T)
```


## Assign taxonomy
```{r assignTaxonomy}
taxa <- assignTaxonomy(seqtab.nochim.collapse, params$db_train, multithread = params$cpus, tryRC = TRUE)
```


```{r taxaPrint}
taxa.print <- taxa  # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

## Export stuff

```{r export_fasta}
#modified from Astrobiomike's tutorial
# giving our seq headers more manageable names (ASV_1, ASV_2...)

master_table <- t(seqtab.nochim.collapse) %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "asv_seqs") %>% 
  left_join(rownames_to_column(as.data.frame(taxa), var = "asv_seqs"), by = "asv_seqs") %>% 
  dplyr::rename(Domain = Kingdom) %>% 
  dplyr::relocate( c("Domain", "Phylum", "Class", "Order", "Family", "Genus"), .before = "asv_seqs") %>% 
  relocate("asv_seqs", .after = last_col()) %>% 
  rownames_to_column(var = "num") %>% 
  mutate(ASV = paste0(paste("ASV", num, sep="_"))) %>% 
  select(-num)
#asv_seqs <- rownames(master_table)
#asv_codes <- vector(dim(master_table)[1], mode="character")


# making and writing out a fasta of our final ASV seqs:
asv_fasta <- master_table %>%
  select(ASV, asv_seqs) %>% 
  dplyr::rename(`seq.name` = ASV,
                `seq.text` = asv_seqs)
phylotools::dat2fasta(asv_fasta, "ASVs.fa")
```


```{r export_counts}
# tax table:
# creating table of taxonomy and setting any that are unclassified as "NA"
ranks <- c("Domain", "Phylum", "Class", "Order", "Family", "Genus")
asv_tax <- master_table %>%
  select(ASV, Domain:Genus)
#row.names(asv_tax) <- sub(">", "", asv_codes)
#colnames(asv_tax) <- ranks
#rownames(asv_tax) <- gsub(pattern = ">", replacement = "", x = asv_codes)

# count table:
asv_tab <- master_table %>% 
  dplyr::select(!c(Domain:Genus, asv_seqs)) %>% 
  dplyr::select(ASV, everything())
#asv_tab <- sapply(as.data.frame(master_table[,(length(ranks)+1):dim(master_table)[2]]), as.numeric)
#row.names(asv_tab) <- sub(">", "", asv_codes)
write.table(asv_tab, "ASVs_counts.tsv", sep = "\t", quote = FALSE, row.names = F)
```


```{r export_tax}
if(params$isits != TRUE) {
  asv_tax_fix <-
    asv_tax %>%
    dplyr::mutate(Genus = ifelse(!is.na(Genus), Genus,
                                 ifelse(!is.na(Family), paste("unclassified", Family, sep = "_"),
                                        ifelse(!is.na(Order), paste("unclassified", Order, sep = "_"),
                                               ifelse(!is.na(Class), paste("unclassified", Class, sep = "_"),
                                                      ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                                             paste("unclassified", Domain, sep = "_")))))),
                  Family = ifelse(!is.na(Family), Family,
                                  ifelse(!is.na(Order), paste("unclassified", Order, sep = "_"),
                                         ifelse(!is.na(Class), paste("unclassified", Class, sep = "_"),
                                                ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                                       paste("unclassified", Domain, sep = "_"))))),
                  Order = ifelse(!is.na(Order), Order,
                                 ifelse(!is.na(Class), paste("unclassified", Class, sep = "_"),
                                        ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                               paste("unclassified", Domain, sep = "_")))),
                  Class = ifelse(!is.na(Class), Class,
                                 ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                        paste("unclassified", Domain, sep = "_"))),
                  Phylum = ifelse(!is.na(Phylum), Phylum,
                                  paste("unclassified", Domain, sep = "_")))
  
  write.table(asv_tax_fix, "ASVs_taxonomy.tsv", sep = "\t", quote=F, row.names = F)
} else if(params$isits == TRUE) {
  asv_tax_fix <-
    as.data.frame(asv_tax) %>%
    dplyr::mutate(
      Genus = ifelse(!is.na(Genus), Genus,
                     ifelse(!is.na(Family), paste("unclassified", Family, sep = "_"),
                            ifelse(!is.na(Order), paste("unclassified", Order, sep = "_"),
                                   ifelse(!is.na(Class), paste("unclassified", Class, sep = "_"),
                                          ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                                 paste("unclassified", Kingdom, sep = "_")))))),
      Family = ifelse(!is.na(Family), Family,
                      ifelse(!is.na(Order), paste("unclassified", Order, sep = "_"),
                             ifelse(!is.na(Class), paste("unclassified", Class, sep = "_"),
                                    ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                           paste("unclassified", Kingdom, sep = "_"))))),
      Order = ifelse(!is.na(Order), Order,
                     ifelse(!is.na(Class), paste("unclassified", Class, sep = "_"),
                            ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                                   paste("unclassified", Kingdom, sep = "_")))),
      Class = ifelse(!is.na(Class), Class,
                     ifelse(!is.na(Phylum), paste("unclassified", Phylum, sep = "_"),
                            paste("unclassified", Kingdom, sep = "_"))),
      Phylum = ifelse(!is.na(Phylum), Phylum,
                      paste("unclassified", Kingdom, sep = "_")))
  
  write.table(asv_tax_fix, "ASVs_taxonomy.tsv", sep = "\t", quote=F, col.names=NA)
}
```


```{r tax_x_counts}
asv_tax_counts <- full_join(asv_tax_fix, asv_tab, by = "ASV")

write.table(asv_tax_counts, "ASVs_taxonomy_x_counts.tsv", sep = "\t", quote=F, row.names = F)
```


```{r phyloseq_relab}
require(phyloseq)

physeq <- phyloseq(otu_table(t(asv_tab %>% column_to_rownames(var = "ASV")), taxa_are_rows = F),
                   tax_table(as.matrix(asv_tax_fix %>% column_to_rownames(var = "ASV"))))

physeq.ra <- transform_sample_counts(physeq, function(x) x / sum(x) )


asv_tax_relab <- cbind(asv_tax_fix, t(otu_table(physeq.ra)))

write.table(asv_tax_relab, "ASVs_taxonomy_x_relab.tsv", sep = "\t", quote=F, row.names = F)
```


```{r phyloseq_export}
saveRDS(physeq, "physeq.RDS")
```

```{r finalRData}
save.image("final.RData")
```

